Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging Patterns of Queen Bumble Bees

### 2023-01-16 script/manuscript_MEE.Rmd
some writing on sdms in discussion

The inability to reliably identify plants down to species can limit our understanding of ecosystem function and interactions (@bortolus2008error). This is especially true for genera where species are defined based upon ecological and behavioral rather than morphological properties, and hence can serve as key habitat bioindicators (e.g. different species of Sagebrush- *Artemisia* L., Willows - *Salix* L., and Sedges - *Carex* L.) (@Gage2013HistoricalRO). The lack of species level data can hinder our understanding of the breadth of habitat which some species occupy, and the interactions they have with other species. Current methods to ameliorate this situation include: ignoring these ecologically relevant levels of detail, revisiting plots as diagnostic material becomes temporally available, assistance from taxonomic specialists, or the use of barcoding or other molecular techniques.

  The identification to species is often mired by lack of diagnostic characters (e.g. flowers, fruits, roots or combinations thereof), an increasing lack of taxonomic experts (@hebert2003biological) and increasingly the description of cryptic species (@janzen2017nuclear, @oliver2009cryptic). And revisiting field sites to identify material using morphological or chemical approaches, can be resource intensive and often does not work.

  Recently barcoding (the identification of a sample from a single organism *e.g.* a piece of leaf), and metabarcoding (the identification of a sample containing a mix of organisms *e.g.* soil), have shown considerable promise in all Kingdoms of life (@ruppert2019past). With plants the identification of members of certain clades using barcoding has been quite successful (@kress2017plant), whereas with other clades results have been elusive (@liu2014identification, @china2011comparative, @coissac2012bioinformatic), however metabarcoding incurs additional challenges to those which  exist for the currently implemented barcodes (@li2015plant, @kress2007two, @cbol2009dna, @coissac2012bioinformatic). Particular challenges with barcodes include the utilization of high-copy number sequences are associated with their rates of divergence, gene tree conflict, and hybridization (@coissac2016barcodes, @fazekas2009plant).

  Currently the largest plant systematic endeavor ever undertaken, the Kew Plant and Fungal Tree of Life (PAFTOL), is approaching completion (@baker2021PAFTOL). This data set will contain hybridization capture (Hyb-Seq) data from at least one species representing each genus in the plant kingdom using the popular A353 probes (@baker2021PAFTOL, @johnson2019universal), resulting in over 14,000 represented species. These publicly available data serve to provide a taxonomically comprehensive backbone for plant metabarcoding, and the A353 probes are currently being used in many other plant phylogenetic issues increasing the sampling depth of many clades (@baker2021exploring). Data from the 10kP project, which seeks to develop reference genomes from a phylogenetically diverse suite of plants will contribute many more records upon it’s intended completion, now slated to be by 2030, similar projects which seek to sequence high amounts of genomes in regions e.g. the ‘Darwin Tree of Life’ are being undertaken which will contribute data  applicable to enormous spatial domains (@cheng2018tenkp, @darwin2022project, @lewin2022biogenome). These data will promote the ability to apply metabarcoding to resolve a diversity of questions relevant to theoretical and applied ecology (@kress2017plant, @hollingsworth2016telling). However, the application of metabarcoding still face challenges relating to the enormity of the genomic data sets and the computational power required to process sequence data. Herein we have resolved major components of the problems of identifying plant material without diagnostic morphological character states using the Angiosperms353 (A353) Hyb-Seq probes (@johnson2019universal), and custom species sequence databases derived via species distribution modelling, and temporal filtering.

  To increase the quality of metabarcoding results in plants, we suggest reducing the number of possible plant species candidates by generating user selected sequence databases relevant to the the study region and its ecological characteristics. To achieve this goal, we first create a list of candidate species using digital collections gleaned from herbaria, survey work, and citizen science (e.g. iNaturalist), from a region exceeding the study area. To these candidate species, modelling approaches - such as logistic regression, may be used to identify taxa which warrant further exploration e.g. modelling to determine their possibility of presence in metabarcoding samples. We then use species distribution models to create potential distribution maps for the candidate species to limit the impact of spatial and taxonomic biases in the species list and account for spatial variations in niche availability throughout the study area. Species distribution models (SDM’s) examine the ecological conditions associated with known occurrence of a species to identify where else in the study area might suitable habitats be found. This approach has the additional benefit of greatly reducing the size of a sequence database, which allows for the usage of genomic size data on personal computers. This can also significantly reduce processing time, particularly as as most next-generation sequence data is deposited as raw-sequence reads.

  Considerable amounts of species interactions vary along time (@caradonna2021seeing). For the tropics and subtropics, contrasts in the flowering periods of many plant species, can provide an additional filter for identifying material in many metagenomic samples (@janzen1967synchronization, @Newstrom1994ANC). In temperate regions, pollination interactions also vary temporally (@caradonna2017interaction), however the overall shorter extent of the active growing season in these systems results in the presence of few to any natural breaks in these systems which reduces the utility of these to operate as filters in the post-processing of sequence matches. Nonetheless, we work through a general approach which seems applicable to the tropics and subtropics to utilize the temporal dimension for classifying sequencing results (but see @davis2022new)

  To test these metagenomic and informatics approaches to determine whether the foraging record of Queen Bumble Bee's is consistent across direct observations and the pollen record, an incongruency in several floral visitation networks involving smaller bodied fauna (@barker2021pollen, @zhao2019topology, @alarcon2010congruence). The assessment of the plant species compositions of pollen is a desired results, with several applications, and numerous complications (@pornon2017dna, @bell2017applying, @sickel2015increased, @bell2019quantitative, @suchan2019pollen,  @johnson2021airborne). The two foraging phases of the Queen Bumble Bee life cycle is essential to 1) increase their weight before diapause, 2) increase their ovary weights while establishing their recently found nests, both of these time periods represent potential demographic bottlenecks in bumble bee populations (@sarro2022bumble). Bumblebees are one of the only groups of insects with unequivocal quantitative evidence for numerous populations declines, while simultaneously serving as the most effective pollinators in temperate montane ecosystems (@cameron2020global, @goulson2008decline, @williams1982distribution, @colla2012assessing, @bergman1996micrometeorological, @bingham1998efficient, @grixti2009decline) often represent the most diverse areas in the temperate and oftentimes offer the sole potential refugia for multiple dimensions of biodiversity under climate change, whilst simultaneously experiencing the greatest proportional changes in mean annual temperature (@brito2018climate, @pepin2022climate). An immediate understanding of how to manage previously overlooked keystone insect species, such as bumble bees, is essential if the refugial potential of the temperate mountains are to be utilized while maintaining their current diversity (@loarie2009velocity, @dobrowski2016climate)

@@

Observations and sample collection was conducted at The Rocky Mountain Biological Laboratory (RMBL; 38°57.5” N, 106°59.3” W (WGS 84), 2900 m.a.s.l.), Colorado, USA (APPENDIX 1 for site information). Pollinator observations of Bombus Latreille spp. (Apidae Latreille) were conducted from June - August of 2015 in six study sites characterized by high-montane/subalpine Parkland vegetation communities. Observations of *Bombus* foraging took place for one hour at each field site in three 100m transects, where all flowers were also counted and placed into abundance bins. Corbiculae loads were, non-lethally, collected once from all Queen individuals encountered.

@@

In order to minimise the number of species for which SDM’s were to be generated, BIEN was queried at a distance of up to 100km from our study area and all plant species records were downloaded. In order to emulate the stochasticity of botanical collecting, this data set was bootstrap re-sampled 250 times, with 90% of samples selected, to create a testing data set. The median of the logistic regression assessing the probability of occurrence of a species record as a function of distance from the study area was used as a threshold distance, under which, to include species as candidates for distribution modelling.

@@

We then generated 4,029 absence points , locations where the focal taxon is anticipated missing, through a random stratification of 19% of the land cover in the area and included them in (@aim2019database). To achieve a larger absence data set, we generated 1,000 pseudo-absence records for each taxon by randomly selecting coordinates located at least 10km away from an occurrence record. For ML models, these pseudo-absences were reduced so that the ratio of presence to absence records were balanced (@barbet2012selecting). To achieve this, we removed absence records inside of 10% of the mean sample value of the presence records; the required number of absence records were then randomly sampled.

We used 26 environmental variables at 30m resolution to predict the potential distribution of each species, six related to climate, five soil, four topographic, four related to cloud cover, with the remaining reflecting assorted abiotic parameters (@wilson2016remotely, @wang2016locally, @hengl2017soilgrids250m, @robinson2014earthenv) (*APPENDIX 6*). **These publicly available data sets, were selected as they ...** . For linear regression models these predictors underwent both *vifstep* (theta = 10, max observations = 12,500) and *vifcor* (theta = 0.7, max observations = 12,500) to detect highly correlated variables, and collinear features were removed leaving 16 variables (@usdm2014).

Modelling: Random Forest and Boosted Regression Trees, were sub sampled with 30% test and two replicates each before weighted ensemble based on True Skill Statistics (tss) (@sdmPackage). Generalised linear models (GLM) and Generalised additive models (GAM) with 30% sub sampling and three replicates each were also ensembled using the tss (@sdmPackage, @). TSS was chosen as the ensemble criterion as it has been shown to work across a range of species occurrences prevalence (@allouche2006assessing). The results of these models were extracted on a cell-by-cell basis to a polygon feature derived from a minimum-spanning tree which encompasses the study sites, and species from either ensemble with greater than 50% mean habitat suitability across all cells were considered present for further purposes (@prim1957minimum).

535 species were modelled using Generalized Linear Models and Generalized Additive Models. 534 species were modelled using Random Forest and Boosted Regression Trees.  To evaluate the accuracy of the species distribution models, additional presence records from GBIF (n = 61,789), and AIM (n = 12,730) were used as test and training sets (n = 74,519) for logistic regression (@gbifDL2021sdms, @aim2019database). Additional novel absence records were generated from the AIM data set to create a data set where each species has balanced presence and absences. 11 or more paired presence and absence records were required for this testing, resulting in 334 species being included in the logistic regression (Mdn = 110.0, $\bar{x}$ = 223.1, max = 1568 record pairs used) with a 70% test split (@caret).

@@

Using five years of observational data on *Bombus* Queen Bee foraging at these studies sites, we identified the plant taxa most frequently visited by Queens across all years. We sequenced the 12 most commonly visited taxa twice using samples from one site within the Gunnison River Drainage and one individual from another population. In addition, for any of these 12 focal species which did not have a congener pair in this filtered sample, we included a congener - or a species from a closely related genus to serve as an outgroup. We also sequenced another 15 abundant taxa commonly visited by *Bombus* workers, based on the aforementioned data set (*APPENDIX 4*). Plant collections were identified via a variety, and typically a combination, of dichotomous keys and primary literature as required (@flora1993flora, @hitchcock2018flora,  @ackerfield2015flora, @lesica2012manual,  @cronquist1977intermountain, @allred2012flora, @jepson2022online, @mohlenbrock2002vascular).

@@

To precipitate the DNA, chilled Isopropyl alcohol & 3 mM Sodium acetate (5:1) equivalent to 2/3 of the volume of sample were added, with 1 hour of chilling at -20°C, followed by 10 minutes of centrifuging at 13,000 rpm. The supernatant was pipetted to a new 1.5 ml centrifuge tube, and 70% EtOH (400 µL) were added before chilling at -20°C for 20 minutes followed by centrifugation at 13,000 rpm for 10 minutes. Both tubes were then washed with  75% EtOH (400 µL), inverted, centrifuged at 13,000 rpm for 4 minutes, and the solution discarded, then washed with 95% EtOH (400 µL) , inverted, centrifuged at 13,000 rpm for 4 minutes, and the solution discarded. Pellets were dried at room temperature overnight before resuspension in Nuclease free H2O.  Extractions were assessed using a Nanodrop 2000 (Thermo Fisher Scientific) and Qubit fluorometer (Thermo Fisher Scientific). DNA extracts were then cleaned using 2:1 v./v. Sera-Mag beads (Cytiva, Little Chalfont, UK) to solute following the manufacturer's protocol, eluted in 0.5x TE, and the eluent allowed to reduce by half volume in ambient conditions. DNA was quantified using a Qubit fluorometer.

@@

Library preparation was performed using the NEBNext Ultra II FS-DNA Library Prep Kit for Illumina (New England BioLabs, Ipswich, Massachusetts, USA) using slightly modified manufacturers recommendation.  Fragmentation was performed at ½ volume of reagents and ¼ enzyme mix for 40 minutes at 37*C, with an input of 500 ng cleaned DNA. Adapter Ligation and PCR enrichment were performed with ½ volumes, while cleanup of products was performed with ½ volume of SPRI beads (Beckman Coulter, Indianapolis, Indiana, USA) and recommended volumes of 80% v./v. ethanol washes. The exception was the herbarium specimens which were not fragmented and only end repaired, with similar library preparation of all samples. Products were analysed on 4% agarose gels, and a Qubit fluorometer. 
Libraries were pooled and enriched with the Angiosperms 353 probe kit V.4 (Arbor Biosciences myBaits Target Sequence Capture Kit) by following the manufacturer's protocol and Brewer et al. 2019. Sequencing was performed using an Illumina mi-Seq with 150-bp end reads, (NUSeq Core, Chicago, Illinois).

@@

A custom Kraken2 database was created by downloading representative species of each genus indicated as being present in the study area by the spatial analyses from the Sequence Read Archive (SRA) NCBI (@wood2019improved). These sequences were processed in the same manner as our novel sequences . The Kraken2 database was built using default parameters. Kraken2 was run on sequences using default parameters (*APPENDIX 5*). 
  Following Kraken2, Bracken was used to classify sequences to terminal taxa (@lu2017bracken). Results from both Kraken2 and Bracken, results were reclassified manually to identify terminal taxa. For example, when only a single species of a genus was known in the study area, but our database used a representative of another taxon in the genus, this species was coded as the result. The re-coding of sequences from another representative species for the genus to the sole RMBL representative allowed the identification of XX & % more species.

@@

A local NCBI database was built using the same processed novel and downloaded sequences as the previous databases (@camacho2009blast).

@@

To develop a reference library of  pollen grains which may be present in corbiculae loads, an image reference collection of fuchsin-jelly stained (@beattie1971technique) slides was assembled from slides previously prepared by the authors (n = 21), and other researchers (n = 38) (@brosi2013single). Using five years of observational data on *Bombus* Queen Bee foraging at these studies sites (Ogilvie unpublished), as well as the Vascular Plant Checklist (@fraser2007vpc), an additional 62 voucher slides for species were prepared and imaged at 400x (Leica DMLB, Leica MC170 HD Camera, Leica Application Suite V. 4.13.0) from non accessioned herbarium collections to supplement the number of species and clades covered (Appendix 3).  
  We used Divisive Hierarchical Clustering techniques to determine which plant taxa were distinguishable via light microscopy, and to develop a dichotomous key to pollen morphotypes. Ten readily discernible categorical traits were collected from each specimen in the image collection. These traits were transformed using Gower distances, and clustered using Divisive Hierarchical clustering techniques (@cluster2022). Using the cluster dendrogram, elbow plot, and heatmaps (@fpc2022), of these results morphological groups of pollen which could not be resolved via microscopy were delineated, and a dichotomous key was prepared (APPENDIX NO.). This key was then used to identify the pollen grains sampled from corbiculae loads to morphotypes in a consistent manner. 
  To prepare the pollen slides from corbiculae, all corbiculae loads were broken apart and rolled using dissection needlepoints to increase heterogeneity of samples. *Cerca* 0.5mm^2 of pollen was placed onto a ~4mm^2 fuchsin jelly cube (@beattie1971technique) atop a graticulated microscope slide, with 20 transects and 20 rows (400 quadrants) (EMS, Hartfield, PA). The jelly was melted, with stirring, until pollen grains were homogeneously spread across the microscope slide. Slides were sealed with Canada Balsam (Rublev Colours, Willits, CA) followed by sealing with nail polish; all samples are noted in *APPENDIX 3*.
  To identify the pollen present in corbiculae loads, light microscopy at 400x (Zeiss Axioscope A1) was used. In initial sampling in three transects, each pollen grain was identified to morphotype and counted; an additional two transects were scanned for morphotypes unique to that slide, if either transect contained an unique morphotype than all grains in that transect were also identified and counted. Subsequent to the first round osf sampling, non-parametric species richness rarefaction curves (@vegans2022), and non-parametric species diversity rarefaction curves were used to assess the completeness of sampling (@inextArticle, @inextPackage). Slides not approaching the asymptote of the rarefaction curve were then re-sampled, and analysed iteratively for up to a total of seven transects *APPENDIX 2*.

@@

To estimate the duration of dates in which plant species were flowering weibull estimates of several phenological parameters all spatially modelled taxa were developed (@belitz2020accuracy, @pearse2017statistical). Only BIEN records which occurred in the Omernik Level 4 Ecoregions within 15km of the study area (n = 5 Level 4 Ecoregions, or conditionally 6 if enough records not be found in the nearest 5), and which were from herbarium records were included. To remove temporally irrelevant herbarium records, i.e. material collected during times which flowering is impossible at the study area due to snow cover, we used the SnowUS data set (@iler2021conceptual, @tran2019cloud) from 2000-2017 was analyzed for the first three days of contiguous snow absence, and the first three days of contiguous snow cover in Fall. Herbarium records after the 3rd quantile for melt, and the 1st quantile for snow cover of these metrics were removed. Species with > 10 records had their weibull distributions generated for the date when 10% of individuals had begun flowering, when 50% were flowering, and when 90% of individuals had flowered.

@@

The median (25.009 km) of the logistic regression assessing the probability of occurrence of a species record as a function of distance from the study area was used as a threshold distance to include species for distribution modelling. A 2-sample test for equality of proportions with continuity correction (X-squared = 13.254, df = 1, p-value = 0.000136, 95% CI 0.04-1.00) was used to test whether more of the records located in the broad ecological sites present at the field station, between the distance of the median (25.009 km) to the third quantile (ca 43.830 km) of the regression distance, where true presences at the field station. Including these records would have resulted in modelling an additional 222 species distributions of which 30 are true presences these taxa were not modelled.

>>>>

The inability to reliably identify plants down to terminal taxon can limit our understanding of ecosystem function and interactions (@bortolus2008error). This is especially true for genera where many species are defined upon ecological and behavioral rather than morphological properties, and hence often serve as bioindicators of habitat (e.g. different species of Sagebrush- *Artemisia* L., Willows - *Salix* L., and Sedges - *Carex* L.) (@Gage2013HistoricalRO). The lack of species level data can hinder our understanding of the breadth of habitat which some species occupy, and the interactions they have with other species. Current methods to ameliorate this situation include: ignoring these ecologically relevant levels of detail, revisiting plots as diagnostic material becomes temporally available, assistance from taxonomic specialists, or the use of barcoding or other molecular techniques (CITE). The identification of organisms to terminal taxon is often mired by lack of diagnostic characters (e.g. flowers, fruits, roots or combinations thereof), an increasing lack of taxonomic experts (@hebert2003biological) and increasingly the description of cryptic species (@janzen2017nuclear, @oliver2009cryptic). And revisiting field sites to identify material using morphological or chemical approaches, can be resource intensive and often does not work.

  Recently barcoding (the identification of a sample from a single organism *e.g.* a piece of leaf), and metabarcoding (the identification of a sample containing a mix of organisms *e.g.* soil), have shown considerable promise in all Kingdoms of Life (@ruppert2019past). With plants the identification of members of certain clades using barcoding has been quite successful (@kress2017plant), whereas with many other clades results have been elusive (@liu2014identification, @china2011comparative, @coissac2012bioinformatic), however metabarcoding incurs additional challenges to those which  exist for the currently implemented barcodes (@li2015plant, @kress2007two, @cbol2009dna, @coissac2012bioinformatic). Particular challenges with the high copy number barcodes (e.g. ITS2, *rbcL*, *matK*, *trnH-psbA*) include the utilization their rates of divergence, gene tree conflict, and hybridization (@coissac2016barcodes, @fazekas2009plant).

  Currently the largest plant systematic endeavor ever undertaken, the Plant and Fungal Tree of Life (PAFTOL) undertaken by the Royal Botanic Gardens Kew, is approaching completion (@baker2021PAFTOL). This data set will contain hybridization capture (Hyb-Seq) data from at least one species in each genus of the plant kingdom using the popular Angiosperms353 (A353) probes, at 353 single-copy orthologous loci, (@baker2021PAFTOL, @johnson2019universal), resulting in over 14,000 represented species. These publicly available data serve to provide a taxonomically comprehensive backbone for plant metabarcoding, and the A353 probes are currently being used in many other plant phylogenetic issues increasing the sampling depth of many clades (@baker2021exploring). Data from the 10kP project, which seeks to develop reference genomes from a phylogenetically diverse suite of plants will contribute many more species upon it’s intended completion, now slated to be by 2030, similar projects which seek to sequence high amounts of genomes in geographic regions e.g. the ‘Darwin Tree of Life’ which will sequence all described taxa in Britain and Ireland, are being undertaken which will contribute data sets applicable to enormous spatial domains (@cheng2018tenkp, @darwin2022project, @lewin2022biogenome). These data will promote the ability to apply metabarcoding to resolve a diversity of questions relevant to theoretical and applied ecology (@kress2017plant, @hollingsworth2016telling). However, the application of metabarcoding still face challenges relating to the enormity of the genomic data sets and the computational power required to process sequence data. Herein we have resolved major components of the problems of identifying plant material without diagnostic morphological character states using the A353 Hyb-Seq probes (@johnson2019universal), and custom species sequence databases derived via species distribution modelling, and temporal filtering.

  To increase the quality of metabarcoding results in plants, we suggest reducing the number of possible plant species candidates by generating user selected sequence databases relevant to the the region of study and its ecological characteristics. To achieve this goal, we first create a list of candidate species using digital collections gleaned from herbaria, survey work, and citizen science (e.g. iNaturalist), from a region exceeding the study area. To these candidate species, modelling approaches - such as logistic regression, may be used to identify taxa which warrant further exploration e.g. modelling to determine their possibility of presence in metabarcoding samples. We then use species distribution models to create potential distribution maps for the candidate species to limit the impact of spatial and taxonomic biases in the species list and account for spatial variations in niche availability throughout the study area. Species distribution models (SDM’s) examine the ecological conditions associated with known occurrence of a species to identify where else in the study area might suitable habitats be found. This approach has the additional benefit of greatly reducing the size of a sequence database, which allows for the usage of genomic size data on personal computers. This approach can also significantly reduce processing time, particularly as as most next-generation sequence data is deposited as raw-sequence reads.

  Considerable amounts of species interactions vary both in space and time (@caradonna2021seeing). In many regions of the world, contrasts in the flowering periods of many plant species, can provide an additional filter for identifying material in certain types of metagenomic samples (@janzen1967synchronization, @Newstrom1994ANC). In high elevation temperate regions, pollination interactions vary temporally and are characterized by high turnover in active periods of species (@caradonna2017interaction), however the overall shorter extent of the active growing season in these systems results in the presence of few to any natural breaks,  which reduces the utility of these to operate as filters in the post-processing of sequence matches. Nonetheless, we work develop a general approach which seems applicable to many areas which utilize the temporal dimension for classifying sequences in metagenomic samples (but see @davis2022new).

  We test these metagenomic and informatics approaches to determine whether the foraging record of Queen Bumble Bee's are consistent across direct observations and the pollen record, an incongruency noted in several floral visitation networks involving smaller bodied fauna (@barker2021pollen, @zhao2019topology, @alarcon2010congruence). The assessment of the plant species compositions of pollen is a desired result, with several applications, and numerous complications (@pornon2017dna, @bell2017applying, @sickel2015increased, @bell2019quantitative, @suchan2019pollen,  @johnson2021airborne). The two foraging phases of the Queen Bumble Bee life cycle is essential to 1) increase their weight before diapause, 2) increase their ovary weights while establishing their recently found nests, both of these time periods represent potential demographic bottlenecks in bumble bee populations (@sarro2022bumble, ...). Bumblebees are one of the only groups of insects with unequivocal quantitative evidence for numerous population declines, while simultaneously serving as the most effective pollinators in many temperate montane ecosystems (@cameron2020global, @goulson2008decline, @williams1982distribution, @colla2012assessing, @bergman1996micrometeorological, @bingham1998efficient, @grixti2009decline) often represent the most diverse areas in the temperate and oftentimes offer the sole potential refugia for multiple dimensions of biodiversity under climate change, whilst simultaneously experiencing the greatest proportional changes in mean annual temperature (@brito2018climate, @pepin2022climate). An immediate understanding of how to manage previously overlooked keystone insect species, such as bumble bees, is essential if the refugial potential of the temperate mountains are to be incorporated into their current diversity (@loarie2009velocity, @dobrowski2016climate).

@@

Observations and sample collection was conducted at The Rocky Mountain Biological Laboratory (RMBL; 38°57.5” N, 106°59.3” W (WGS 84), 2900 m.a.s.l.), Colorado, USA (APPENDIX 1 for site information). Pollinator observations of *Bombus* Latreille spp. (Apidae Latreille) were conducted from June - August of 2015 in six study sites characterized by high-montane/subalpine Parkland vegetation communities. Observations of *Bombus* foraging took place for one hour at each field site in three 100m transects, where all flowers were also counted and placed into abundance bins. Corbiculae loads were, non-lethally, collected once from all Queen individuals encountered.

@@

In order to minimise the number of species for which SDM’s were to be generated, BIEN was queried at a distance of up to 100km from our study area and all plant species records were downloaded. In order to emulate the stochasticity of botanical collecting and offset the number of records associated with the research station, this data set was bootstrap re-sampled 250 times, with 90% of samples selected, to create a testing data set. The median of the logistic regression assessing the probability of occurrence of a species record as a function of distance from the study area was used as a threshold distance, under which, to include species as candidates for distribution modelling.

@@

We then generated 4,029 absence points, locations where the focal taxon is anticipated missing, through a random stratification of 19% of the land cover in the area and included them in (@aim2019database). To achieve a larger absence data set, we generated 1,000 pseudo-absence records for each taxon by randomly selecting coordinates located at least 10km away from an occurrence record. For ML models, these pseudo-absences were reduced so that the ratio of presence to absence records were balanced (@barbet2012selecting). To achieve this, we removed absence records inside of 10% of the mean sample value of any predictor variable the presence records; the required number of absence records were then randomly sampled.

To predict the potential distribution of each species we used 26 environmental variables at 30m resolution, six related to climate, five soil, four topographic, four related to cloud cover, with the remaining reflecting assorted abiotic parameters (@wilson2016remotely, @wang2016locally, @hengl2017soilgrids250m, @robinson2014earthenv) (*APPENDIX 6*). **These publicly available data sets, were selected as they ...** . For linear regression models these predictors underwent both *vifstep* (theta = 10, max observations = 12,500) and *vifcor* (theta = 0.7, max observations = 12,500) to detect highly correlated variables, and collinear features were removed leaving 16 variables (@usdm2014).

Modelling: Random Forest and Boosted Regression Trees, were sub sampled with 30% test and two replicates each before weighted ensemble based on True Skill Statistics (tss) (@sdmPackage). Generalised linear models (GLM) and Generalised additive models (GAM) with 30% sub sampling and three replicates each were also ensembled using the tss (@sdmPackage, @). TSS was chosen as the ensemble criterion as it has been shown to work across a wide range of species occurrences prevalence (@allouche2006assessing). The results of these models were extracted on a cell-by-cell basis to a polygon feature derived from a minimum-spanning tree which encompasses the study sites, and species from either ensemble with greater than 50% mean habitat suitability across all cells were considered present for further purposes (@prim1957minimum).

535 species were modelled using Generalized Linear Models and Generalized Additive Models. 534 species were modelled using Random Forest and Boosted Regression Trees.  To evaluate the accuracy of the species distribution models, additional presence records from GBIF (n = 61,789), and AIM (n = 12,730) were used as test and training sets (n = 74,519) for logistic regression (@gbifDL2021sdms, @aim2019database). Additional novel absence records were generated from the AIM data set to create a data set where each species has balanced presence and absences. Eleven or more paired presence and absence records were required for this testing, resulting in 334 species being included in the logistic regression (Mdn = 110.0, $\bar{x}$ = 223.1, max = 1568 record pairs used) with a 70% test split (@caret).

@@

Using five years (2015-2020) of observational data on *Bombus* Queen Bee foraging at these studies sites, we identified the plant taxa most frequently visited by Queens across all years. We sequenced the 12 most commonly visited taxa twice using samples from one site within the Gunnison Basin River Drainage and one individual from another more distal population. In addition, for any of these 12 focal species which did not have a congener pair in this filtered sample, we included a congener - or a species from a closely related genus to serve as an outgroup. We also sequenced another 15 abundant taxa commonly visited by *Bombus* workers, based on the abundances, and immediate access to plant tissue, in the aforementioned data set (*APPENDIX 4*). Plant collections were identified via a variety, and typically a combination, of dichotomous keys and primary literature as required (@flora1993flora, @hitchcock2018flora,  @ackerfield2015flora, @lesica2012manual,  @cronquist1977intermountain, @allred2012flora, @jepson2022online, @mohlenbrock2002vascular).

@@

To precipitate the DNA, chilled Isopropyl alcohol & 3 mM Sodium acetate (5:1) equivalent to $\frac{2}{3}$ of the volume of sample were added, with 1 hour of chilling at -20°C, followed by 10 minutes of centrifuging at 13,000 rpm. The supernatant was pipetted to a new 1.5 ml centrifuge tube, and 70% EtOH (400 µL) were added before chilling at -20°C for 20 minutes followed by centrifugation at 13,000 rpm for 10 minutes. Both tubes were then washed with  75% EtOH (400 µL), inverted, centrifuged at 13,000 rpm for 4 minutes, and the solution discarded, then washed with 95% EtOH (400 µL) , inverted, centrifuged at 13,000 rpm for 4 minutes, and the solution discarded. Pellets were dried at room temperature overnight before resuspension in nuclease free H~2~O.  Extractions were assessed using a Nanodrop 2000 (Thermo Fisher Scientific) and Qubit fluorometer (Thermo Fisher Scientific). DNA extracts were then cleaned using 2:1 v./v. Sera-Mag beads (Cytiva, Little Chalfont, UK) to solute ratio following the manufacturer's protocol, eluted in 0.5x TE, and the eluent allowed to reduce by half volume in ambient conditions. DNA was quantified using a Qubit fluorometer.

@@

Library preparation was performed using the NEBNext Ultra II FS-DNA Library Prep Kit for Illumina (New England BioLabs, Ipswich, Massachusetts, USA) using slightly modified manufacturers recommendation.  Fragmentation was performed at ½ volume of reagents and ¼ enzyme mix for 40 minutes at 37°C, with an input of 500 ng cleaned DNA. Adapter Ligation and PCR enrichment were performed with ½ volumes, while cleanup of products was performed with ½ volume of SPRI beads (Beckman Coulter, Indianapolis, Indiana, USA) and recommended volumes of 80% v./v. ethanol washes. The exception was the herbarium specimens which were not fragmented and only end repaired, with similar library preparation of all samples. Products were analysed on 4% agarose gels, and a Qubit fluorometer. 
Libraries were pooled and enriched with the Angiosperms 353 probe kit V.4 (Arbor Biosciences myBaits Target Sequence Capture Kit) by following the manufacturer's protocol and Brewer et al. 2019. Sequencing was performed using an Illumina mi-Seq with 150-bp end reads, (NUSeq Core, Chicago, Illinois).

@@

A custom Kraken2 database was created by downloading representative species of each genus indicated as being present in the study area by the spatial analyses from the Sequence Read Archive (SRA) NCBI (@wood2019improved). These sequences were processed in the same manner as our novel sequences. The Kraken2 database was built using default parameters. Kraken2 was run on sequences using default parameters (*APPENDIX 5*). 
  Following Kraken2, Bracken was used to classify sequences to terminal taxa (@lu2017bracken). Results from both Kraken2 and Bracken, results were reclassified manually to identify terminal taxa. For example, when only a single species of a genus was known in the study area, but our database used a representative of another taxon in the genus, this species was coded as the result. The re-coding of sequences from another representative species for the genus to the sole RMBL representative allowed the identification of *XX & %* more species.

@@

A local NCBI database was built using the same processed novel and downloaded sequences as the previous database (@camacho2009blast).

@@

To develop a reference library of  pollen grains which may be present in corbiculae loads, an image reference collection of fuchsin-jelly stained (@beattie1971technique) slides was assembled from slides previously prepared by the authors (n = 21), and other researchers (n = 38) (@brosi2013single). Using five years of observational data on *Bombus* Queen Bee foraging at these studies sites (Ogilvie unpublished), as well as the Vascular Plant Checklist (@fraser2007vpc), an additional 62 voucher slides for species were prepared and imaged at 400x (Leica DMLB, Leica MC170 HD Camera, Leica Application Suite V. 4.13.0) from non accessioned herbarium collections to supplement the number of species and clades covered (Appendix 3).  
  We used Divisive Hierarchical Clustering techniques to determine which plant taxa were distinguishable via light microscopy, and to develop a dichotomous key to pollen morphotypes. Ten readily discernible categorical traits were collected from each specimen in the image collection. These traits were transformed using Gower distances, and clustered using Divisive Hierarchical clustering techniques (@cluster2022). Using the cluster dendrogram, elbow plot, and heatmaps (@fpc2022), of these results morphological groups of pollen which could not be resolved via microscopy were delineated, and a dichotomous key was prepared (APPENDIX NO.). This key was then used to identify the pollen grains sampled from corbiculae loads to morphotypes in a consistent manner. 
  To prepare the pollen slides from corbiculae, all corbiculae loads were broken apart and rolled using dissection needlepoints to increase heterogeneity of samples. *Cerca* 0.5mm^2^ of pollen was placed onto a ~4mm^2^ fuchsin jelly cube (@beattie1971technique) atop a graticulated microscope slide, with 20 transects and 20 rows (400 quadrants) (EMS, Hartfield, PA). The jelly was melted, with stirring, until pollen grains were homogeneously spread across the microscope slide. Slides were sealed with Canada Balsam (Rublev Colours, Willits, CA) followed by sealing with nail polish to prevent oxidation; all samples are noted in *APPENDIX 3*.
  To identify the pollen present in corbiculae loads, light microscopy at 400x (Zeiss Axioscope A1) was used. In initial sampling in three transects, each pollen grain was identified to morphotype and counted; an additional two transects were scanned for morphotypes unique to that slide, if either transect contained an unique morphotype than all grains in that transect were also identified and counted. Subsequent to the first round of sampling, non-parametric species richness rarefaction curves (@vegans2022), and non-parametric species diversity rarefaction curves were used to assess the completeness of sampling (@inextArticle, @inextPackage). Slides not approaching the asymptote of the rarefaction curve were then re-sampled, and analysed iteratively for up to a total of seven transects *APPENDIX 2*.

@@

To estimate the duration of dates in which plant species were flowering weibull estimates of several phenological parameters all spatially modelled taxa were developed (@belitz2020accuracy, @pearse2017statistical). Only BIEN records which occurred in the Omernik Level 4 Ecoregions within 15km of the study area (n = 5 Level 4 Ecoregions, or conditionally 6 ecoregions if enough records were not be found in the nearest 5), and which were from herbarium records were included. To remove temporally irrelevant herbarium records, i.e. material collected during times which flowering is impossible at the study area due to snow cover, we used the SnowUS data set (@iler2021conceptual, @tran2019cloud) from 2000-2017 were analyzed for the first three days of contiguous snow absence, and the first three days of contiguous snow cover in Fall. Herbarium records after the 3^rd^ quantile for melt, and the 1^st^ quantile for snow cover of these metrics were removed. Species with > 10 records had their weibull distributions generated for the date when 10% of individuals had begun flowering, when 50% were flowering, and when 90% of individuals had flowered, we used the initiation and cessation dates, respectively, as effective start and ends of flowering.

@@

The median (25.009 km) of the logistic regression assessing the probability of occurrence of a species record as a function of distance from the study area was used as a threshold distance to include species for distribution modelling. A 2-sample test for equality of proportions with continuity correction (X-squared = 13.254, df = 1, p-value = 0.000136, 95% CI 0.04-1.00) was used to test whether more of the records located in the broad ecological sites present at the field station, between the distance of the median (25.009 km) to the third quantile (ca 43.830 km) of the regression distance, were true presences at the field station. Including these records would have resulted in modelling an additional 222 species distributions of which 30 are true presences, these taxa were not modelled.

Across the entire spatial domain of modelling all ensembled models (n = 968) had an accuracy of 0.84 (95% CI 0.8356 - 0.8443), kappa = 0.68, p-value < 0.001, sensitivity = 0.80, specificity = 0.87, AUC = 0.92.
